# Preface {.unnumbered}

This book is about doing microeconometrics with R. Microeconometrics is defined by @CAME:TRIV:05\index[author]{Cameron}\index[author]{Trivedi} as "the analysis of individual-level data on the economic behavior of individuals or ﬁrms using regression methods applied to cross-section and panel data". We'll use in this book a broader definition of microeconometrics as some of the empirical examples use countries or regions as a unit of observation. But, nevertheless, the underlying model will have microeconomic foundations, as for example the Solow growth model that is treated at length in @sec-mult_reg_chapter and @sec-spatial_econometrics and the Kuzets curve presented in @sec-interpretation_chapter. A negative delimitation of what is in this book is that it doesn't contain any analyzes of time series. 

I used throughout the book some manuals. My main reference was @CAME:TRIV:05, but I also used @GREE:18 (especially the online mathematical and statistical appendices that I found particularly useful) and @WOOL:10. At an introductory level, I also used @STOC:WATS:15 and @MADD:LAHI:09. For some points (geometry of least squares, asymptotic theory, data generating process, computational considerations), I was also inspired by Davidson and MacKinnon [-@DAVI:MACK:93; -@DAVI:MACK:04]\index[author]{Davidson}\index[author]{MacKinnon}.
On more specific subjects, I used also some specialized textbook and surveys: @MADD:83\index[author]{Maddala}, Amemiya [-@AMEM:81; -@AMEM:84]\index[author]{Amemiya} for @sec-binomial and @sec-tobit (binomial and censored models), @CAME:TRIV:13\index[author]{Cameron}\index[author]{Trivedi} for @sec-count (count data),
@LANC:90\index[author]{Lancaster} and @KIEF:88\index[author]{Kiefer} for @sec-duration (duration models) and @TRAI:09\index[author]{Train} for @sec-rum (discrete choice models).

## R and RStudio {-}

**R** appeared in the late nineties as a clone of **S** which rapidly became a **GNU** project. It became increasingly popular among statisticians, especially in fields where **S+** was widely used. Its large adoption by econometricians is more recent, especially because of the use by econometricians of other softwares like **SAS**, **GAUSS** and especially nowadays **Stata**, and the lack of some popular estimators and tests for econometrics in **R**. Since then, numerous excellent packages have been developed that make the adoption of **R** as the primary software for an econometrician a relevant choice. Moreover, **R** has some characteristics that make **R** code concise and clear, compared to other softwares. 

**Rstudio** (now **Posit**), founded in 2009 by Joseph J. Allaire\index[author]{Allaire}, developed an integrated development environment  called **Rstudio** and a set of **R** packages that changed deeply the way a lot of **R** users (including myself) use **R**. Among them, **quarto** was used to build this book and the set of packages called the **tidyverse** is used extensively for all the basic tasks of an applied statistician which are, citing @WICK:CETI:GROL:23\index[author]{Wickham}\index[author]{Cetinkaya-Rundel}\index[author]{Grolemund}, importing, tidying, transforming and visualizing data sets. Packages of the **tidyverse**, respectively **readr**, **tidyr**, **dplyr** and **ggplot2** enable to perform these tasks using a set of efficient, intuitive and consistent functions. 

This book requires basic knowledge of **R** and the **tidyverse**, at the level of the first part of @WICK:CETI:GROL:23, which is untitled "the whole game" and is an excellent introduction for readers of this book that are not **R** users. Moreover, there is a free on-line version of this book that can be found at <https://r4ds.hadley.nz/>.

## R packages

**R** has a modular structure. A particular analysis with **R** requires the use of functions that are part of different **packages**. To use these functions, the user can either "attach" the package (using the `library` function) or prefix the name of the function with the name of the package. For example, to use the `waldtest` function of the **lmtest** package, one can either use:

```{r}
#| eval: false
library(lmtest)
waldtest(---)
```

or:

```{r}
#| eval: false
lmtest::waldtest(---)
```

A relevant strategy is to attach only the packages that are often used in the analysis.

Different kind of packages should be considered:

- the **base** packages contain the core of **R**. They are included in any distribution of **R** and they are automatically attached, which means that the user can access directly to the functions they contain. The two most important are **base** and **stats**, which contains respectively the basic general and statistical functions of **R**. 
- the **R** distribution also contains 15 **recommended** packages, which are not attached. Among them, we'll use in this book **survival** and **MASS**.
- **contributed** R packages. They are developed by **R** users and are hosted by the Comprehensive R Archive Network (**CRAN**). They can be very easily installed within the Rstudio IDE (using the `Packages` tab on the right down panel of Rstudio). Nowadays, there are more than 20000 of them. 

20 years ago, using **R** for econometrics analysis required a lot of programming because a lot of core methods of econometrics were not available neither in the core of **R**, nor in contributed packages. Nowadays, most of the basic methods are available in contributed packages and the book uses extensively these packages. Some missing methods are implemented in the companion package of the book called **micsr**.

## Data sets {-}

This book uses a lot of data sets to illustrate the use of the techniques described in the book. A particular care was given to use original and interesting data sets. Our main source to get data was the Journal of Applied Econometrics data archive maintained by James G. MacKinnon\index[author]{MacKinnon} since 1994 and available at <http://qed.econ.queensu.ca/jae/>^[This website is still available but is superseded since 2022 by a new website at <https://journaldata.zbw.eu/journals/jae>.] and the website of the American Economic Association <https://www.aeaweb.org> which provides data sets and programs used in articles published in AEA's journal, in particular the American Economic Review and the four American Economic Journals (we particularly used Applied Economics). We also wrote to a lot of authors to ask them whether the data used in their publications was still available and if they could share it with us. Many thanks for those who provided data sets that are now included in the **micsr** and the **micsr.data** packages: John Mullahy\index[author]{Mullahy} (`birthwt`, `cigmales`), Mark Ottoni Wilhelm\index[author]{Wilhem} (`charitable`), Lawrence Katz\index[author]{Katz} (`recall`), Joseph Terza\index[author]{Terza} (`trips`), Franck Koppelman\index[author]{Koppelman} (`toronto_montreal`), Christopher Skeels\index[author]{Skeels} and David Drukker\index[author]{Moffitt} (`moffitt`), Winfried Pohlmeier\index[author]{Pohlmeier} (`doctor_ger`), Rainer Winkelman\index[author]{Winkelman} (`jobmob`).^[Franck Vella\index[author]{Vella}, Adrian Pagan\index[author]{Pagan}, Jerry Hausman\index[author]{Hausman}, Claudia Goldin\index[author]{Goldin}, Wynand van de Ven\index[author]{van de Ven}, Andrew Jones\index[author]{Jones}, Richard J Smith\index[author]{Smith}, Pramila Krishnan\index[author]{Krishnan}, Orley C. Ashenfelter\index[author]{Ashenfelter}, John Ham\index[author]{Ham}, James Adams\index[author]{Adams}, John McDonald\index[author]{McDonald}, Laurence Kotlikoff\index[author]{Kotlikoff}, Mark Rosenzweig\index[author]{Rosenzweig}, James Dunvely\index[author]{Dunvely} and Kathryn Graddy\index[author]{Graddy} kindly answered me that they were unable to share the data I required (most of the time because it was not available anymore to them).]

## The micsr package {-}

The book comes with a companion package called **micsr**. It is available on **CRAN** and is therefore very easy to install. As indicated previously, the book use extensively the **tidyverse** package and it is therefore recommended to install **tidyverse** before trying to replicate the **R** code contained in the book. **micsr** contains a small subset of the data sets included in the book and some specific methods of estimations and testing procedures for microeconometrics. Data sets used in this book not included in the **micsr** package are available in the **micsr.data** package which is available from **github**. To install it, you should first install the **pak** package and then simply write in the console:

```{r}
#| eval: false
pak::pkg_install("ycroissant/micsr.data")
```

Throughout the book we assume that the reader that is willing to replicate the examples had previously attached the **tidyverse**, **micsr** and **micsr.data** packages, using:

```{r}
library(tidyverse)
library(micsr)
library(micsr.data)
```

**micsr** provides specific estimation and testing methods that will be presented in the core of the book. It also provides some general purpose functions. Among them, the `gaze` function will be extensively used throughout the book.

The output of **R** calls is often quite long and takes two much place in a book. For this reason, **micsr** contains a generic function called `gaze` with methods for a lot of **R** objects. Consider for example a t-test:

```{r}
t.test(extra ~ 1, data = sleep)
```

The output is 12 lines long, including 3 blank lines. `gaze` writes the main part of the results in just 1 line:

```{r}
#| collapse: true
t.test(extra ~ 1, data = sleep) %>% gaze
```

<!-- **micsr** also contains some general utilities particularly useful to extract some information from fitted models. Consider the following linear model: -->

<!-- ```{r} -->
<!-- z <- lm(dist ~ speed, cars) -->
<!-- summary(z) -->
<!-- ``` -->

<!-- The printed output of `summary` is quite long and contains informations that can't be easily extracted from the result. A `gaze` method for `lm` object is provided to print only the table of coefficients (by default without the intercept): -->

<!-- ```{r} -->
<!-- #| collapse: true -->
<!-- z %>% gaze -->
<!-- ``` -->

<!-- With base **R**, its easy to extract the number of observations using `nobs`, but there is no such function to extract the number of parameters. We provide the `npar` function for this purpose: -->

<!-- ```{r} -->
<!-- #| collapse: true -->
<!-- npar(z) -->
<!-- ``` -->

<!-- The second column of the table of the coefficients is the vector of standard errors of the coefficients With base **R**, it can be extract by first using the `vcov` function to extract the covariance matrix and then by taking the square roots of the diagonal elements of this matrix. **micsr** provide the `stder` function to achieve easily the same task: -->

<!-- ```{r} -->
<!-- #| collapse: true -->
<!-- z %>% vcov %>% diag %>% sqrt -->
<!-- z %>% stder -->
<!-- ``` -->

<!-- Two very important statistics of the linear model are the coefficient of determination (adujsted or not) and the F statistic. They should be extracted "manually" from the object returned by `summary`, applied to a `lm` object: -->

<!-- ```{r} -->
<!-- #| collapse: true -->
<!-- summary(z)$r.squared -->
<!-- summary(z)$adj.r.squared -->
<!-- summary(z)$fstat -->
<!-- ``` -->

<!-- The `rsq` and the `ftest` of the **micsr** package can be used instead: -->

<!-- ```{r} -->
<!-- #| collapse: true -->
<!-- z %>% rsq -->
<!-- z %>% rsq(type = "adj") -->
<!-- z %>% ftest %>% gaze -->
<!-- ``` -->

\mainmatter
